---
title: "Using R Packages to Disseminate Data"
author: "Chris Hansen"
date: "2016-07-20"
archives: "2016"
categories: ["R", "shiny", "R packages"]
tags: ["R", "shiny", "R packages"]
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
   comment = NA, echo = TRUE, cache = FALSE,
   fig.align = "center",
   warning = FALSE, message = FALSE
)
```
```{r imports, include = FALSE, echo = FALSE}
library(data.table)
library(NZCensus)
library(NZCensusDemos)
```

# Introduction

R packages can include datasets, and it is very common to do so.  This shouldn't
really come as a surprise--R is a statistical language in the end, and it is
normal to include working examples with pacakges; and including datasets against
which to run unit tests and other examples is clearly a useful thing to do.

But datasets themselves aren't central to most packages--statistical 
methodologies often are, and the data is there for testing or illustrative
purposes.  But this doesn't have to be the case.  For analysts familiar with R,
consuming packages to obtain data might be quite convenient.  But we can also
bundle functionality specific to the included datasets, possibly increasing its
overall utility.  For example, one could include pre-canned statistical analyses
in the form of simple helper functions or interactive `shiny` applications.

In this document, we look at a couple of datasets available from the
[Statistics New Zealand](http://www.stats.govt.nz)
website and consider the ways in which value might be added by disseminating
this data via an R package (or set of packages) along with bundled demo apps.


# Anatomy of an R Package

An R package is essentially an eponymously named folder which
bundles additional functionality in a standardised way, and which can be loaded
on demand. A package contains, among other things, function and type
definitions, documentation, and datasets.

Traditionally, a loadable version of a package would be built from a source 
repository by running shell commands such as:

```
R CMD INSTALL --build pkgname
```

or

```
R CMD build pkgname 
R CMD install pkgname.tar.gz
```

depending on the target platform.  It is not all that difficult to build a
modest package, but it can be tedious to construct packages from scratch.
Large packages, on the other hand, can be difficult to build and manage.  Those
interested in the detail should refer to:

[Writing R Extensions](https://cran.r-project.org/)


## Building packages in RStudio

RStudio has a number of productivity features which greatly assist the authoring
of packages.  RStudio allows users to work with projects, and packages can be
authored by creating a project with type 'R Package':

<img src=assets/package01.png class="large" alt="Project->New Project...">
<img src=assets/package02.png class="large" alt="Create Project">
<img src=assets/package03.png class="large" alt="Project Type">
<img src=assets/package04.png class="large" alt="Create R Package">

When working on a package, RStudio provides a number of useful tools via a
'Build' menu.  From there you can build the package and create source or binary
packages for dissemination.

<img src=assets/package05.png class="large" alt="Build menu">

In addition to this, help files can be built automatically from source code 
comments using the `roxygen2` package.  The traditional method is to make a
markup file for each function that requires documentation, but using `roxygen2`
is far more convenient.  We'll see examples below, but for a reasonable overview
see:

[Object documentation](http://r-pkgs.had.co.nz/man.html)

Finally, a package source code can be managed via either of git or subversion,
which brings obvious advantage.

<img src=assets/versioning.png class="large" alt="Version control">


# Single-function Shiny Applications

It is possible to make 1-page web applications in R via the `shiny` package.
There are a number of ways a `shiny` application can be created, and RStudio
tries to help out here as well.  

For example, selecting File &rarr; New File &rarr; Shiny Web App... in RStudio 
will cause RStudio to create a small working app which you can use as a starting
point:


<img src=assets/shiny01.png class="large" alt="Shiny web app">
<img src=assets/shiny02.png class="large" alt="New Shiny Web Application">
<img src=assets/shiny03.png class="large" alt="Shiny application running within RStudio IDE">

A `shiny` application is essentially made of a `ui` component which contains the
UI definition and a `server` component which contains all the server-side logic
for the application.  Given these components, there are then a number of ways of
creating and running an application.  For example, one could create files
`foo/ui.R` and `foo/server.R` which contain the `ui` and `server` components,
respectively, and then run the application by calling: 

```r
shiny::runApp("foo")
```

Alternatively, if `ui` and `server` are actual R objects, one could run an
application via

```r
shiny::runApp(list(ui=ui, server=server))
```

or even 

```r
shiny::shinyApp(ui=ui, server=server)
```

But, of course, we can then write a function which returns one of these
application objects.  For example

```r
foo <- function(){
   shinyApp(ui=ui, server=server)
}
```

so that an application can be started, then, simply by calling:

```
foo()
```

In fact, this is what we will do here.  We will include functions in the
packages we create which do nothing but cause a `shiny` application to
run.  Thus, we can provide users with data, but also interactive applications
that make use of that data.


# Example--Census Meshblock Dataset

For our first example, let us consider the Census Meshblock Dataset.  From the
[Statistics New Zealand website](http://www.stats.govt.nz/Census/2013-census/data-tables/meshblock-dataset.aspx): 

> 2013 Census meshblock dataset contains counts at meshblock and other
geographic levels for selected variables from the 2013, 2006, and 2001 Censuses.
The geographies correspond to 2013 Census boundaries. The counts are at the
highest level of each variable’s classification.

> The meshblock dataset is in Excel (.xlxs) and CSV format and can be integrated 
with analytical or mapping systems (using geographic boundary files), enabling 
further analysis of census data.

We'll look specifically at the
[downloadable CSV file](http://www3.stats.govt.nz/meshblock/2013/csv/2013_mb_dataset_Total_New_Zealand_CSV.zip),
primarily because the CSV file would be of the most universal use to any kind of
data analyst (casual users might use the Excel version, but even those users 
might find it more useful to use the R version we'll create here).

## Repairing / reorganising the data

This data has a number of organisational issues that impedes its use, so I spent
a bit of time tidying things up and put the results in a package called
[NZCensus](https://github.com/cmhh/NZCensus).

## Making a package

Here we make a simple package which does nothing but provide the following five
datasets:

* `datadictionary`
* `dwelling`
* `household`
* `family`
* `individual`

That is, once installed, users can refer to any of these datasets by name if
the package is loaded, or else load any of the datasets into memory by calling:

```r
data(dataset, package="NZCensus")
```

Note that if all 5 datasets were loaded into memory, the total footprint would
be 318MB.  However, the package itself weighs in at 40.3MB--this compares
favourably with the original CSV source (a zip file), which weighs in at 37.6MB.

We use RStudio's integrated support for packages to make this as easy as
possible.  We create a project called `NZCensus`, which will also become
the name of the package.  The basic layout is as follows:

```asis
NZCensus/data/datadictionary.rda
NZCensus/data/dwelling.rda
NZCensus/data/family.rda
NZCensus/data/household.rda
NZCensus/data/individual.rda
NZCensus/inst/doc/datadictionary.xlsx
NZCensus/inst/doc/datadictionary.ods
NZCensus/inst/doc/datadictionary.csv
NZCensus/man/datadictionary.Rd
NZCensus/man/dwelling.Rd
NZCensus/man/family.Rd
NZCensus/man/housheold.Rd
NZCensus/man/individual.Rd
NZCensus/R/data.R
NZCensus/DESCRIPTION
NZCensus/NAMESPACE
```

The `*.rda` files are datasets in R's native format, and are compressed.  For
example, the individual dataset is 209.8MB uncompressed in-memory.  But this can
be compressed and saved to disk by running:

```r
save(individual, file = "individual", compress = TRUE)
```

with the resulting file in this case being just 28.7MB.

R packages are usually very well documented.  Here, we take advantage of
`doxygen2` to build help files automatically from the package R scripts.
In this case, we don't actually have any R code, but we fake it anyway.  We
create entries for each dataset in a file called `./NZCensus/R/data.R`.  For
example, for the `dwelling` dataset we include the following:

```r
#' New Zealand Census of Population and Dwellings dwelling data
#'
#' A dataset containing dwelling counts at various geographic levels.
#'
#' @format A data table with columns:
#' \describe{
#'   \item{geography}{Geographic level.  One of MB (meshblock), AU (area unit),
#'   Ward, TA (Territorial Authority), or RC (Regional Council).}
#'   \item{code}{Geography code.}
#'   \item{description}{Geography description.}
#'   \item{year}{Census year.  One of 2001, 2006, or 2013.}
#'   Other variables are documented in \code{\link{datadictionary}}.
#' }
#' @source
#' \url{http://www.stats.govt.nz/Census/2013-census/data-tables/meshblock-dataset.aspx}
"dwelling"
```

Processing this file yields, among other things, `NZCensus/man/dwelling.Rd`,
which is in R's own documentation format.  And after actually building and
installing this package, an HTML version of this file is created which can
be queried from R via something like:

```r
?NZCensus::dwelling
```

<img src=assets/dwellinghelp.png class=large alt="Help for dwelling dataset.">

Besides this, RStudio also automates the creation of other files, but several
files were also created manually.  For example, files were added to the folder
`NZCensus/inst/doc/`, and once the package has been installed, these are
available from the main help page via the
`User guides, package vignettes and other documentation.` link.

<img src=assets/NZCensushelp.png class=large alt="help(package='NZCensus')">
<img src=assets/NZCensusdoc.png class=large alt="NZCensus docs">

In this case, these files are just versions of the `datadictionary`
dataset that can be opened in Excel or LibreOffice (what used to be known as
OpenOffice).


## Extending the package

Above, we wrapped five datasets up as a package which we called `NZCensus`.  We
can add value to packages by bundling other interesting functions along with the
data.  For example, we could write a function that takes a particular variable
name as input, and plots that variable for each Census year.

In this case, we'll create two functions that provide `shiny` applications
in the way described above:

* `mapdemo` - an interactive map which displays a choropleth based on user
  input
* `distributiondemo` - an barplot showing the distribution of a 
  categorical variable selected by the user.

We place these functions in a new package called
[`NZCensusDemos`](https://github.com/cmhh/NZCensusDemos) which has an explicit
dependency on the `NZCensus` package, among others.  The main reason for this is
that we will bundle some additional spatial feature classes required to draw
geographical features such as regional council areas on a map.  The feature
classes aren't terribly large (we haven't included, amusingly, given the product
name, the meshblock feature class, which is pretty large), but at least this way
users who only want the Census datasets aren't burdened with the extra bulk
implied by the demo package if they don't want it.  Similarly, the demos have
dependencies which would need to be installed along with them, and loaded any
time the demo package is.  So, again, if users have no need for the demos, then
they wouldn't be forced to install these additional dependencies if they didn't
want them.

Either way, the `mapdemo` function, heavily edited, looks as follows:

```r
mapdemo <- function(){
   shinyApp(
      ui = fluidPage(
         titlePanel("Mapping Census..."),
         sidebarLayout(
            sidebarPanel(width=3,
               ...
            ),
            mainPanel(width=9,
               ...
            )
         )
      ),
      server = function(input, output){
         ...
      }
   )
}
```

It can then be called via:

```r
library(NZCensusDemos)
mapdemo()
```

and looks as follows:

<img src=assets/NZCensusMapDemo.png class=large alt="NZCensusDemos::mapdemo()">

Similarly, the `distributiondemo` can be called via:

```r
distributiondemo()
```

and looks as follows:

<img src=assets/NZCensusDistributionDemo.png class=large alt="NZCensusDemos::distributiondemo()">


# Example--HLFS Time Series Data

Since looking at the Census meshblock dataset had convinced myself that putting
data in R pacakges was a useful thing to do, I started thinking about other
datasets for which this sort of treatment might also be useful.

[Statistics New Zealand](http://www.stats.govt.nz) puts bulk CSV files on their
website containing a number of time series, and that seemed a reasonable choice.
We can use this data to produce a `shiny` application which lets users explore a
range time series and make selections to plot.


## Preparing the data sets

The data I decided to use is labour market data found via the
[CSV files for download](http://www.stats.govt.nz/tools_and_services/releases_csv_files.aspx)
page here:

[Income and work](http://www.stats.govt.nz/~/media/Statistics/Browse%20for%20stats/LabourMarketStatistics/HOTPMar16qtr/lms-mar16-zipped-csv-tables.zip)

The data consists of 14 columns, and 605535 rows.
There isn't any major problems with this data--it's perfectly usable as it is.
However there is a fair bit of duplication:

<img src=assets/rawlabmkt.png class=large alt="lots of duplication!">

Specifically, the `subject`, `group`, and title columns are unique by
`series_reference`.  So, we split the table into two tables which have
`series_reference` as a common key, and remove as much duplication as possible.
In addtion, we take the opportunity to cast the variable `Period` to a genuine
date-type variable.    

<img src=assets/metadata.png class=large alt="metadata">
<img src=assets/labmkt.png class=large alt="metadata">


## Making a package

In this case, the data is relatively small, and for convenience we bundle the
data and demos together.  We create a package called
[`SNZlabmkt`](https://github.com/cmhh/SNZlabmkt) which contains the two data
sets, `metadata` and `labmkt`, as well as a single function, `timeseriesdemo`,
which provides a `shiny` application which plots time series based on user
input.  Since it has been discussed above, we skip the details of package
creation here.  But the demo is triggered by calling:

```r
library(SNZlabmkt)
timeseriesdemo()
```

and it looks as follows:

<img src=assets/SNZlabmktTSDemo.png class=large alt="timeseriesdemo()">


# Try it out...

The packages discussed here have all been uploaded to GitHub.  RStudio has
excellent integration with Git, as noted earlier, and allows users to use GitHub
to set a remote repository.

<img src=assets/git01.png class=large alt="Git">
<img src=assets/git02.png class=large alt="Git">
<img src=assets/git03.png class=large alt="Git">

To install the packages:

```r
if (!require(devtools)) install.packages("devtools")
devtools::install_github("cmhh/NZCensus")
devtools::install_github("cmhh/NZCensusDemos")
devtools::install_github("cmhh/SNZlabmkt")
```

If for some reason this fails, download each file manually through from:

* [NZCensus](https://github.com/cmhh/NZCensus/archive/master.zip)
* [NZCensusDemos](https://github.com/cmhh/NZCensusDemos/archive/master.zip)
* [SNZlabmkt](https://github.com/cmhh/SNZlabmkt/archive/master.zip)

Then install each zip file in turn as follows:

```r
devtools::install_local("path-to-zip/NZCensus-master.zip")
devtools::install_local("path-to-zip/NZCensusDemos-master.zip")
devtools::install_local("path-to-zip/SNZlabmkt-master.zip")
```


